---
title: "SAR flood detectability"
subtitle: "When no Sentinel-1 pass falls near the flood peak: is the event date separable from the rest of the season?"
author: "Manish Bilore"
date: today
params:
  event: mumbai_2024
---

```{r setup}
source("R/helpers.R")
ev <- params$event
P  <- ev_paths(paste0(ev, "_detectability"))
S  <- fromJSON(file.path(P$out, "summary.json"))
ts <- tb(P, "water_timeseries"); det <- tb(P, "detectability")
D  <- file.path(ROOT, "data/events", ev)
edate <- as.Date(S$event_date)
fmt <- function(x, d = 1) ifelse(is.na(x), "–", format(round(x, d), big.mark = ","))
```

::: {.callout-important appearance="simple"}
**`r gsub("_", " ", ev)` — no acquisition near the flood peak.**
Relative orbit `r S$path` is the only track over this area, at a `r S$repeat_days`-day repeat: `r S$n_passes` passes in the season.
The pass under test is **`r S$event_date`**. Verdict: the event date is **`r toupper(S$verdict)`** from the rest of the season.
:::

# Why this event gets a different report

Every other event in this project had a Sentinel-1 acquisition inside the flood peak, so a map of that scene *is* a flood map.
Here it does not. Urban pluvial flooding in Mumbai drains within hours of the rain stopping, while the satellite returns every
`r S$repeat_days` days on a single track — so the question is not "how much flooding does the map show" but **"could this
system have seen the flood at all?"**

The test: apply the identical classical rule to **every** pass of the monsoon, and compare the event date against the others.
If the flood left a detectable signature, the event date should stand out from the seasonal and tidal spread. If it does not,
the extent measured on that date is indistinguishable from an ordinary wet-season scene — which is the answer, and a limit of
the observing system rather than of the method.

# Acquisitions

```{r acq, fig.height=1.9, fig.width=10}
dd <- data.frame(date = as.Date(unlist(S$dates)))
dd$kind <- ifelse(dd$date == edate, "event pass", "other pass")
ggplot(dd, aes(date, 1, colour = kind, shape = kind)) + geom_point(size = 3.4) +
  scale_colour_manual(values = c(`event pass` = "#b2182b", `other pass` = "grey55"), name = NULL) +
  scale_shape_manual(values = c(`event pass` = 17, `other pass` = 16), name = NULL) +
  scale_x_date(date_breaks = "1 month", date_labels = "%d %b") +
  labs(x = NULL, y = NULL, title = sprintf("Every Sentinel-1 pass over the AOI (path %s, %d-day repeat)", S$path, S$repeat_days)) +
  theme_rep() + theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank())
```

# Water extent across the season

```{r series, fig.height=4.2, fig.width=10}
if (!is.null(ts)) {
  ts$date <- as.Date(ts$date)
  d <- ts %>% filter(method %in% c("otsu_vv", "split_vv", "cd_vv")) %>% mutate(method = mlab(method))
  ggplot(d, aes(date, area_km2, colour = method)) +
    geom_line(linewidth = .5) + geom_point(size = 1.8) +
    geom_vline(xintercept = edate, linetype = 2, colour = "#b2182b") +
    annotate("text", x = edate, y = max(d$area_km2), label = " event pass", colour = "#b2182b", hjust = 0, size = 3) +
    scale_x_date(date_breaks = "1 month", date_labels = "%d %b") +
    labs(x = NULL, y = "mapped water extent [km²]", colour = NULL,
         title = "The same rule applied to every pass of the season") + theme_rep()
} else na_note("Time series", "python src/detectability.py <event>")
```

```{r series_builtup, fig.height=3.4, fig.width=10}
if (!is.null(ts) && "builtup_km2" %in% names(ts)) {
  d <- ts %>% filter(method %in% c("otsu_vv", "cd_vv", "urban_incr")) %>% mutate(method = mlab(method))
  ggplot(d, aes(date, builtup_km2, colour = method)) + geom_line(linewidth = .5) + geom_point(size = 1.8) +
    geom_vline(xintercept = edate, linetype = 2, colour = "#b2182b") +
    scale_x_date(date_breaks = "1 month", date_labels = "%d %b") +
    labs(x = NULL, y = "mapped water on built-up land [km²]", colour = NULL,
         title = "Built-up fraction — where urban flooding would show up") + theme_rep()
}
```

# Is the event date separable?

```{r det_tab, results='asis'}
if (!is.null(det)) {
  print(kbl(det %>% transmute(method = mlab(method), `event km²` = event_km2, `season mean km²` = others_mean_km2,
                              `season sd` = others_sd_km2, `season max km²` = others_max_km2,
                              `z-score` = z_score, `rank in season` = rank_of_season,
                              `min detectable km²` = min_detectable_km2, separable), 2))
  s <- det %>% filter(method %in% c("otsu_vv", "split_vv", "cd_vv")) %>% rename(season_sd = others_sd_km2)
  cat(sprintf(paste0(
    "\n**z-score** is how many seasonal standard deviations the event pass sits above the mean of the other passes; ",
    "**rank** is its position among all %d passes (1 = largest extent of the season). A flood that the sensor caught should ",
    "score z ≥ 2 and rank 1 on every sound method.\n\n"), S$n_passes))
  if (S$verdict == "not separable") {
    cat(sprintf(paste0("Here it does neither: the classical methods put the event pass at z = %s and rank %s of %d. ",
      "The extent mapped on %s is **inside the ordinary spread of the season** — what any single scene shows in this AOI is ",
      "dominated by permanently wet ground, creeks and antecedent rainfall rather than by the flood. Whatever water the event ",
      "left on the surface was either gone by the overpass or too small and too mixed with buildings to register at 10 m.\n\n"),
      paste(sprintf("%.2f", s$z_score), collapse = " / "), paste(s$rank_of_season, collapse = " / "), S$n_passes, S$event_date))
    cat(sprintf(paste0("The same table sets a **detection floor** for this AOI and sampling. Season-to-season scatter is %.1f–%.1f km² ",
      "(1 sd), so a flood has to add **%.0f–%.0f km² of newly mapped water** before it clears z = 2 — against a seasonal mean of ",
      "%.0f–%.0f km². The 11 July pass added %+.1f–%+.1f km². That floor, not the choice of threshold or model, is what decides ",
      "whether an event of this kind is observable at all.\n\n"),
      min(s$season_sd), max(s$season_sd), min(s$min_detectable_km2), max(s$min_detectable_km2),
      min(s$others_mean_km2), max(s$others_mean_km2), min(s$event_excess_km2), max(s$event_excess_km2)))
  } else {
    cat(sprintf(paste0("The event pass scores z = %s and rank %s of %d — the flood signature survives the %d-day sampling. ",
      "Note that the three methods share an input and a threshold family, so they are not independent evidence: treat this as one ",
      "test passed three ways, not three confirmations.\n\n"),
      paste(sprintf("%.2f", s$z_score), collapse = " / "), paste(s$rank_of_season, collapse = " / "), S$n_passes, S$repeat_days))
  }
}
```

# Event pass against its closest rival

```{r maps, fig.height=5, fig.width=11}
agg <- function(f, k = 5) { r <- tryCatch(rast(f), error = function(e) NULL); if (is.null(r)) NULL else aggregate(r, k, "mean", na.rm = TRUE) }
if (!is.null(ts)) {
  cd <- ts %>% filter(method == "cd_vv")
  rival <- cd %>% filter(!is_event) %>% slice_max(area_km2, n = 1)
  f1 <- file.path(D, "baseline", format(edate, "%Y%m%d"), "flood_cd_vv.tif")
  f2 <- file.path(D, "baseline", format(as.Date(rival$date[1]), "%Y%m%d"), "flood_cd_vv.tif")
  mapgrid(list(agg(f1), agg(f2)),
          c(sprintf("event pass %s — %.1f km²", S$event_date, cd$area_km2[cd$is_event][1]),
            sprintf("largest other pass %s — %.1f km²", as.Date(rival$date[1]), rival$area_km2[1])),
          PAL$flood, c(0, 1), ncol = 2, leg = "flooded\nfraction")
}
```

If the two panels look alike, that is the finding: the event pass is another wet-season scene.

# What this means for monitoring

- **Revisit, not method, is the binding constraint for urban pluvial floods.** A `r S$repeat_days`-day repeat on one track
  cannot be expected to catch water that drains in hours. No amount of threshold tuning or deep learning changes that.
- **Geometry matters as much as timing.** Only one relative orbit covers this AOI, so there is no second chance from an
  ascending pass a few hours later on the same day.
- **What would work**: a constellation with sub-daily revisit (Sentinel-1 C + D from 2026, or commercial X-band tasking),
  or a different observable — rain-gauge and drainage-network data for the event itself, with SAR reserved for the
  slower riverine and coastal flooding it does see well.
- The rest of this project shows the method working where an acquisition exists (Kerala, Assam). This page is the honest
  boundary of that claim.

```{r caveats, results='asis'}
cat(sprintf(paste0("\n::: {.callout-note appearance=\"simple\"}\n**Caveats.** The comparison uses one relative orbit (%s) and identical ",
  "processing on every date, so the passes are directly comparable; the thresholds are re-estimated per scene, as they would be ",
  "operationally. Tidal stage is not controlled — the intertidal band is masked, but creeks and mudflats still move between passes, ",
  "which is part of the seasonal spread being measured. A flood too small or too brief to register at 10 m would also produce this ",
  "result; separability is evidence about the observing system, not proof that no water was present.\n:::\n"), S$path))
```
